---
title: "Unsupervised learning - clustering and dimension reduction"
author: "Anna Yeaton"
date: "Fall 2019"
output:
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
```

# Lab Section
Download auto data from the *Statistical Learning* book website here: http://www-bcf.usc.edu/~gareth/ISL/data.html

Today, we are going over Hierarchical clustering, K-Means Clustering, PCA, and ICA. 

```{r load, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ggfortify)
library(fastICA)
```

```{r}
# read in Auto data
Auto_data <- read_csv("/Users/ay1392/Desktop/Auto.csv")

#remove cars with unknown horsepower and set horsepower to numeric
Auto_data <- Auto_data %>% 
  filter(horsepower != "?") %>% 
  mutate(horsepower = as.numeric(horsepower)) %>% 
  as.data.frame()

#save car names 
Auto_data_names <- Auto_data$name

#data to cluster
Auto_data_clust <- Auto_data[,1:8]
dim(Auto_data_clust)

#392 is too much for a demo, so lets take the first 25
Auto_data_clust <- Auto_data_clust[1:25,]
rownames(Auto_data_clust) <- Auto_data_names[1:25]
```


## Hierarchical agglomerative clustering

Step 1. Assign each item to it's own cluster. We start with 25 clusters, one for each car. 

Step 2. Calculate a similarity matrix between each cluster.

Step 3. Find the pair of clusters closest in similarity. 

Step 4. Merge these clusters/recalculate similarity between clusters. Options are: single linkage (nearest neighbor), complete linkage (furthest neighbor), average linkage (mean distance between all pairs of data from the two different clusters), centroid linkage (distance between the means of all points in the clusters). 

Step 5. Repeat Step 3 and 4 until there is only one cluster.

### In practice

Step 1. Each car is a cluster. 

Step 2. Create a distance matrix from Auto_data_clust.

```{r}
help("dist")
hierarchical_dist <- as.matrix(dist(Auto_data_clust, method = "euclidean"))
#View(hierarchical_dist)
```

Step 3. Find the two cars that are the most similar to each other and print the names of those two cars

```{r }
diag(hierarchical_dist) <- NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 23 and 15 are the most similar. Lets go back to the names of the cars
Auto_data_names[23]
Auto_data_names[15]

```

Step 4. Merge the two clusters together using average linkage. 

```{r }
#replace pos 15 with the average of pos 15 and 23
hierarchical_dist[,15] <- apply((hierarchical_dist[,c(23,15)]),1,mean)
hierarchical_dist[15,] <- apply((hierarchical_dist[c(23,15),]),2,mean)

#remove pos 23
hierarchical_dist <- hierarchical_dist[-23,-23]

#now position 15 represents the cluster containing the saab99e and the toyota corona mark ii
```

Step 5. To complete the algorithm, go back to step 3 and iterate through all of the previous steps until there are no more rows left

```{r }
diag(hierarchical_dist) <- NA
arrayInd(which.min(hierarchical_dist), dim(hierarchical_dist))

#postitions 4 and 3 are the most similar
Auto_data_names[4]
Auto_data_names[3]

```

### R function 

Now that we know how the algorithm works, let's use the R function hclust. Plot the Dendogram resulting from clustering the Auto_data_clust using average linkage.  

```{r}
hierarchical_dist <- dist(Auto_data_clust, method = "euclidean")
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
```

There is one more element to hierarchical clustering: Cutting the tree. Here, we can control how many clusters we want or the height of the tree. 
```{r}
#help(cutree)
# cut tree into 3 clusters
tree <- hclust(hierarchical_dist, method="average")
plot(tree)
tree_k2 <- cutree(tree, k = 2)
rect.hclust(tree, k = 3, h = NULL)
```


\newpage

## K-Means Clustering
Step 1. Choose the N number of clusters.

Step 2. Find the N items that are furthest apart and set them as cluster centroids.

Step 3. Assign one item in the dataset to the closest of the N cluster centroids.

Step 4. Recalculate the cluster centroid. 

Step 5. Repeat Steps 3 and 4 until all items are in a cluster. 

Step 6. Go through each item and reassess whether the item belongs in the current cluster or in a different cluster based on distance to cluster centroids. Every time an item is reassigned to a different cluster, the centroids must be recalculated.

Step 7. When every item belongs firmly to a cluster, or the iterations of Step 6 are endless, the algorithm is complete. 

### In practice

Step 1. We are going  to cluster the 25 cars into two groups.

Step 2a. Find the two cars furthest from each other. 

```{r K-means_step1}
kmeans_dist <- as.matrix(dist(Auto_data_clust, method = "euclidean"))
diag(kmeans_dist) <- NA
arrayInd(which.max(kmeans_dist), dim(kmeans_dist))
```

Step 2b. Create data frames to hold each cluster, cluster names, and centroids. 

```{r}
cluster_one <- Auto_data_clust[20,]
cluster_one_names <- Auto_data_names[20]
cluster_one_centroid <- cluster_one

cluster_two <- Auto_data_clust[9,]
cluster_two_names <- Auto_data_names[20]
cluster_two_centroid <- cluster_two
```

Step 3a. Sequentially put cars in either cluster one or cluster two. lets start with car 1. Is car 1 closer to cluster one or cluster two?

```{r}
#distance to cluster 1
dist(rbind(cluster_one_centroid,Auto_data_clust[1,]), method = "euclidean") 

#distance to cluster 2
dist(rbind(cluster_two_centroid,Auto_data_clust[1,]), method = "euclidean")
```

Step 3b. Add car 1 to cluster two and adjust the centroid value for cluster 2. The new centroid value is a mean of the values of cars in that cluster. 
```{r}
cluster_two <- rbind(cluster_two, Auto_data_clust[1,])
cluster_two_names <- rbind(cluster_two_names, Auto_data_names[1])
cluster_two_centroid <- apply(cluster_two,2,function(x) mean(as.numeric(x)))
```

Step 3d. Do for all cars

```{r}
for(i in 1:nrow(Auto_data_clust)){
  if(i == 9 | i == 20){
    next
  }
  if(dist(rbind(cluster_two_centroid,Auto_data_clust[i,]),
          method = "euclidean") < dist(rbind(cluster_one_centroid,
                                             Auto_data_clust[i,]), method = "euclidean") ) {
    cluster_two <- rbind(cluster_two, Auto_data_clust[i,])
    cluster_two_names <- rbind(cluster_two_names, Auto_data_names[i])
    cluster_two_centroid <- apply(cluster_two,2,function(x) mean(as.numeric(x)))
  }else{
    cluster_one <- rbind(cluster_one, Auto_data_clust[i,])
    cluster_one_names <- rbind(cluster_one_names, Auto_data_names[i])
    cluster_one_centroid <- apply(cluster_one,2,function(x) mean(as.numeric(x)))
  }
}
```

Step 4. Adjust the clusters by comparing the distance of each car to the centroid of its current cluster versus the distance to the centroid of the other cluster. Does it still belong in the current cluster? lets start with car 1. Does it belong in cluster two?

```{r}
dist(dist(rbind(cluster_one_centroid,Auto_data_clust[1,])))
dist(dist(rbind(cluster_two_centroid,Auto_data_clust[1,])))
```

Alright, so car 1 does belong in cluster two. To complete the algorithm, iterate over the cars until none of the cars switch clusters.

### R function

Now we know how the algorithm works, lets use the R function kmeans. 

```{r}
kmean_out <- kmeans(Auto_data_clust,2)
kmean_out
```

\newpage

## Principal Components Analysis (PCA)

Principal Components Analysis is a linear dimensionality reduction algorithm. If you want to learn more about linear algebra, I suggest the MIT Open Courseware class here : https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/
There are two ways of doing PCA, Single Value Decomposition (SVD), and the method we will use today, using the covariance matrix of the data. 

Step 1. Center data by subtracting the mean.

Step 2. Calculate covariance matrix of data.

Step 3. Perform Eigendecomposition of the covariance matrix. i.e. represent the matrix in terms of it's eigenvalues and eigen vectors

Step 4. Multiply the eigen vectors by the original data to express the data in terms of the eigen vectors. 

Step 1. Center the data by subtracting the mean of the each column from the values in that column

```{r}
Auto_data_clust_pca <- data.matrix(Auto_data_clust)

Center_auto <- apply(Auto_data_clust_pca, 2, function(x) x - mean(x))
```


Step 2. Calculate covariance matrix of the Auto data

```{r}

Covariance_auto <- cov(Center_auto)
```

Step 3.  Calculate eigen values and vectors

```{r}
Eigen_value_auto <- eigen(Covariance_auto)$value

#columns are the eigen vectors
Eigen_vector_auto <- eigen(Covariance_auto)$vector
```

Step 4. Multiply the eigen vector matrix by the original data. 

```{r}
PC <- as.data.frame(data.matrix(Center_auto) %*% Eigen_vector_auto)

ggplot(PC, aes(PC[,1], PC[,2])) + geom_point(aes(PC[,1], PC[,2]))
#+ geom_text(aes(label=Auto_data_names[1:8]), nudge_x = -2.5, nudge_y = 400)
```

Step 5. Find out which principal components explain the variance in the data. 

```{r}
#for each component, take the cumulative sum of eigen values up to that point and and divide by the total sum of eigen values
round(cumsum(Eigen_value_auto)/sum(Eigen_value_auto) * 100, digits = 2)
```

Principal component 1 and 2 explain 99.99 percent of the variance. Principal component 1,2, and 3 together explain 100% of the variance in the data. 

### R function 
Now that we know how PCA works, lets use the R funtion prcomp.

```{r}
help("prcomp")
autoplot(prcomp(Auto_data_clust_pca))
```

\newpage

## Independent Component Analysis (ICA)
ICA is an algorithm that finds components that are independent, subcomponents of the data. 

Step 1. Whiten the data by projecting the data onto the eigen vectors (PCA).

Step 2. Solve the X=AS equation by maximizing non-gaussianty in the variables(components) in S. 

This results in a matrix S with components that are independent from each other. 

We will use the fastICA algorithm.

First we will go backwards. 
Create a matrix S with the independent components
```{r}
#create two signals
S <- cbind(cos((1:500)/10), ((500:1)/1000))

par(mfcol = c(1, 2))
plot(S[,1], type="l")
plot(S[,2], type="l")
```

Create a mixing matrix A
```{r}
A <- matrix(c(0.5, 0.7, 0.423, 0.857), 2, 2)
```

Mix S using A
```{r}
X <- S %*% A
par(mfcol = c(1, 2))
plot(X[,1], type="l")
plot(X[,2], type="l")

```

Unmix using fastICA
```{r, include=FALSE}
a <- fastICA(X, 2, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)
```


```{r}
par(mfcol = c(1, 2))
plot(1:500, a$S[,1], type = "l", xlab = "S'1", ylab = "")
plot(1:500, a$S[,2], type = "l", xlab = "S'2", ylab = "")
```


### ICA on the auto data
```{r, include=FALSE}
a <- fastICA(Auto_data_clust, 7, alg.typ = "parallel", fun = "logcosh", alpha = 1,
             method = "R", row.norm = FALSE, maxit = 200,
             tol = 0.0001, verbose = TRUE)
```

plot the independent components as a heatmap
```{r}
heatmap(a$S)
```


\newpage

## Homework

```{r}
data(iris)
```

0. Subset the Iris dataset to only include `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width`. 

1. Run PCA on the Iris dataset. Plot a scatter plot of PC1 vs PC2 and include the percent variance those PCs describe. 

2. Run ICA on the Iris dataset. Plot the independent components as a heatmap.

3. Use Kmeans to cluster the Iris data. 
  * Use the silhouette function in the cluster package to find the optimal number of clusters for kmeans for the iris dataset. Then cluster using kmeans clustering. Does the data cluster by species? 
  * Using this clustering, color the PCA plot according to the clusters.
  
4. Use hierarchical clustering to cluster the Iris data.

  * Try two different linkage types, and two different distance metrics. 
  * For one linkage type and one distance metric, try two different cut points. 
  * Using this clustering, color the PCA plot according to the clusters. (6  plots in total)
  

# Optional material
On PCA:

Eigen Vectors and Eigen Values http://www.visiondummy.com/2014/03/eigenvalues-eigenvectors/
Linear Algebra by Prof. Gilbert Strang https://ocw.mit.edu/courses/mathematics/18-06-linear-algebra-spring-2010/video-lectures/
http://www.cs.otago.ac.nz/cosc453/student_tutorials/principal_components.pdf
https://stats.stackexchange.com/questions/2691/making-sense-of-principal-component-analysis-eigenvectors-eigenvalues

On ICA: 

Independent Component Analysis: Algorithms and Applications https://www.cs.helsinki.fi/u/ahyvarin/papers/NN00new.pdf
Tutorial on ICA taken from http://rstudio-pubs-static.s3.amazonaws.com/93614_be30df613b2a4707b3e5a1a62f631d19.html



